---
title: "The Global Digital Divide"
author: "D/ADC/26/0032"
date: "2026-03-24"
output: html_document
---

# 1. Introduction

Every day, billions of people use the internet to work, learn, and connect.
But in 2024, **2.6 billion people have never used the internet.**

This analysis explores the Global Digital Divide — the gap between 
countries with high internet access and those left behind.

**Research Question:**
> "Which countries are most affected by the digital divide, 
>  and does wealth determine internet access?"

---
```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(countrycode)
library(RColorBrewer)
library(plotly)
library(knitr)
```
---

# Load & Clean Data|Load raw datasets
```{r}
internet_raw <- read_csv("C:/Users/User1/Desktop/DigitalDivide/R asm/internet_users.csv", skip = 4)
gdp_raw      <- read_csv("C:/Users/User1/Desktop/DigitalDivide/R asm/gdp_per_capita.csv", skip = 4)
mobile_raw   <- read_csv("C:/Users/User1/Desktop/DigitalDivide/R asm/mobile_subscriptions.csv", skip = 4)
```

---
```{r}
# Clean Internet Users dataset
internet_clean <- internet_raw %>%
  # select only columns we need
  select(`Country Name`, `Country Code`, `2000`:`2022`) %>%
  # remove rows with no country name
  filter(!is.na(`Country Name`)) %>%
  # pivot from wide to long format
  # wide = each year is a column
  # long = one row per country per year
  pivot_longer(
    cols = `2000`:`2022`,
    names_to = "Year",
    values_to = "Internet_Users_Pct"
  ) %>%
  # convert Year from text to number
  mutate(Year = as.integer(Year)) %>%
  # add continent information
  mutate(
    Continent = countrycode(`Country Code`,
                            origin = "iso3c",
                            destination = "continent")
  ) %>%
  # remove rows where internet % is missing
  filter(!is.na(Internet_Users_Pct)) %>%
  # remove regional aggregates (keep only real countries)
  filter(!is.na(Continent))

# Clean GDP dataset (same process)
gdp_clean <- gdp_raw %>%
  select(`Country Name`, `Country Code`, `2000`:`2022`) %>%
  filter(!is.na(`Country Name`)) %>%
  pivot_longer(
    cols = `2000`:`2022`,
    names_to = "Year",
    values_to = "GDP_Per_Capita"
  ) %>%
  mutate(Year = as.integer(Year)) %>%
  filter(!is.na(GDP_Per_Capita))

# Clean Mobile dataset
mobile_clean <- mobile_raw %>%
  select(`Country Name`, `Country Code`, `2000`:`2022`) %>%
  filter(!is.na(`Country Name`)) %>%
  pivot_longer(
    cols = `2000`:`2022`,
    names_to = "Year",
    values_to = "Mobile_Subscriptions"
  ) %>%
  mutate(Year = as.integer(Year)) %>%
  filter(!is.na(Mobile_Subscriptions))

# Merge all 3 datasets together
combined_data <- internet_clean %>%
  left_join(gdp_clean %>%
              select(`Country Code`, Year, GDP_Per_Capita),
            by = c("Country Code", "Year")) %>%
  left_join(mobile_clean %>%
              select(`Country Code`, Year, Mobile_Subscriptions),
            by = c("Country Code", "Year"))

# Save cleaned data for Shiny app to use
write_csv(combined_data, "combined_data.csv")

# Quick check
cat("Total rows:", nrow(combined_data), "\n")
cat("Countries:", n_distinct(combined_data$`Country Name`), "\n")
cat("Years:", min(combined_data$Year), "to", max(combined_data$Year), "\n")
```

---

# 4. Chart 1 — Internet Growth by Continent
```{r}
# Calculate average internet usage per continent per year
continent_trend <- combined_data %>%
  group_by(Continent, Year) %>%
  summarise(
    Avg_Internet = mean(Internet_Users_Pct, na.rm = TRUE),
    .groups = "drop"
  )

# Create the line chart
chart1 <- ggplot(continent_trend,
                 aes(x = Year,
                     y = Avg_Internet,
                     color = Continent,
                     group = Continent)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set2") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Internet Usage Growth by Continent (2000-2022)",
    subtitle = "The digital divide between continents is growing",
    x = "Year",
    y = "Average Internet Users (%)",
    color = "Continent",
    caption = "Source: World Bank"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50"),
    legend.position = "right"
  )

# Make it interactive
ggplotly(chart1)
```

**Insight:** Europe and the Americas have seen rapid growth since 2000,
while Africa remains significantly behind despite recent improvements.

---

# 5. Chart 2 — Top vs Bottom Countries
```{r chart2, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
# Get most recent year data (2022)
latest_data <- combined_data %>%
  filter(Year == 2022) %>%
  group_by(`Country Name`, Continent) %>%
  summarise(
    Internet_Pct = mean(Internet_Users_Pct, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(Internet_Pct))

# Top 10 countries
top10 <- latest_data %>%
  arrange(desc(Internet_Pct)) %>%
  head(10) %>%
  mutate(Group = "Top 10")

# Bottom 10 countries
bottom10 <- latest_data %>%
  arrange(Internet_Pct) %>%
  head(10) %>%
  mutate(Group = "Bottom 10")

# Combine top and bottom
top_bottom <- bind_rows(top10, bottom10)

# Create bar chart
chart2 <- ggplot(top_bottom,
                 aes(x = reorder(`Country Name`, Internet_Pct),
                     y = Internet_Pct,
                     fill = Group)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Top 10" = "#B23AEE",
                                "Bottom 10" = "#FF6A6A")) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "The Digital Divide: Top 10 vs Bottom 10 Countries (2022)",
    subtitle = "A gap of over 90 percentage points separates leaders from laggards",
    x = "Country",
    y = "Internet Users (%)",
    fill = "Group",
    caption = "Source: World Bank"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray50")
  )

ggplotly(chart2)
```

**Insight:** Countries like Norway and Denmark have near 100% internet access,
while countries like South Sudan and Chad remain below 10%.

---

# 6. Chart 3 — Scatter Plot: Wealth vs Internet Access
```{r chart3, message=FALSE, warning=FALSE, fig.width=10, fig.height=7}
# Get 2022 data with both internet and GDP
scatter_data <- combined_data %>%
  filter(Year == 2022) %>%
  filter(!is.na(GDP_Per_Capita)) %>%
  filter(!is.na(Internet_Users_Pct)) %>%
  group_by(`Country Name`, Continent) %>%
  summarise(
    Internet_Pct = mean(Internet_Users_Pct, na.rm = TRUE),
    GDP = mean(GDP_Per_Capita, na.rm = TRUE),
    .groups = "drop"
  )

# Create scatter plot
chart3 <- ggplot(scatter_data,
                 aes(x = GDP,
                     y = Internet_Pct,
                     color = Continent,
                     text = `Country Name`)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "loess",
              se = TRUE,
              color = "black",
              linetype = "dashed") +
  scale_x_log10(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Does Wealth Determine Internet Access?",
    subtitle = "GDP per capita vs Internet usage % (2022)",
    x = "GDP Per Capita (log scale, USD)",
    y = "Internet Users (%)",
    color = "Continent",
    caption = "Source: World Bank"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray50")
  )

ggplotly(chart3, tooltip = c("text", "x", "y"))
```

**Insight:** There is a very strong positive relationship between GDP per capita
and internet access. Wealthier countries consistently have higher internet usage.
This confirms that economic development is a key driver of digital inclusion.

---

# 7. Conclusion

This analysis reveals three key findings:

1. **The divide is real** — Internet access ranges from under 5% to nearly 100%
2. **Continents matter** — Africa lags far behind Europe and Americas
3. **Wealth drives access** — GDP per capita strongly predicts internet usage

The digital divide is not just about technology — it reflects deeper 
inequalities in economic development and infrastructure investment.

> *"In a world where the internet is increasingly essential for education,
> work, and healthcare — being offline means being left behind."

---




